---
title: 'Lab 3: Reserve Planning'
author: "Clarissa Boyajian"
date: "2/6/2022"
output: 
  html_document:
    number_sections: true
    code_folding: hide
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Set up

Load packages and set parameters
```{r}
if (!require("librarian")){
  install.packages("librarian")
  library(librarian)
}
librarian::shelf(assertthat, BiocManager, tidyverse, gridExtra, here, mapview, 
  prioritizr, prioritizrdata, raster, remotes, rgeos, rgdal, scales, sf, sp, 
  units)
if (!require("lpsymphony")){
  BiocManager::install("lpsymphony")
  library(lpsymphony)
}

options(scipen = 999)

redo <- FALSE
```

Data setup
```{r}
dir_data <- here("data/prioritizr")
pu_shp   <- file.path(dir_data, "pu.shp")
pu_url   <- "https://github.com/prioritizr/massey-workshop/raw/main/data.zip"
pu_zip   <- file.path(dir_data, basename(pu_url))
vegetation_tif <- file.path(dir_data, "vegetation.tif")

dir.create(dir_data, showWarnings = F, recursive = T)
if (!file.exists(pu_shp) | redo){
  download.file(pu_url, pu_zip)
  unzip(pu_zip, exdir = dir_data)
  dir_unzip   <- file.path(dir_data, "data")
  files_unzip <- list.files(dir_unzip, full.names = T)
  file.rename(
    files_unzip, 
    files_unzip %>% str_replace("prioritizr/data", "prioritizr"))
  unlink(c(pu_zip, dir_unzip), recursive = T)
}
```


# Data

Data import
```{r}
# import planning unit data
pu_data <- as(read_sf(pu_shp), "Spatial")

# format columns in planning unit data
pu_data$locked_in <- as.logical(pu_data$locked_in)
pu_data$locked_out <- as.logical(pu_data$locked_out)

# import vegetation data
veg_data <- stack(vegetation_tif)
```

## Planning unit data

```{r}
# print a short summary of the data
print(pu_data)

# plot the planning unit data
plot(pu_data)
```

```{r, eval = FALSE}
# plot an interactive map of the planning unit data
mapview(pu_data)
```

```{r}
# print the structure of object
str(pu_data, max.level = 2)
```

```{r}
# print the class of the object
class(pu_data)
```

```{r}
# print the slots of the object
slotNames(pu_data)
```

```{r}
# print the coordinate reference system
print(pu_data@proj4string)
```

```{r}
# print number of planning units (geometries) in the data
nrow(pu_data)
```

```{r}
# print the first six rows in the data
head(pu_data@data)
```

```{r}
# print the first six values in the cost column of the attribute data
head(pu_data$cost)
```

```{r}
# print the highest cost value
max(pu_data$cost)
```

```{r}
# print the smallest cost value
min(pu_data$cost)
```

```{r}
# print average cost value
mean(pu_data$cost)
```

```{r, out.width = "60%"}
# plot a map of the planning unit cost data
spplot(pu_data, "cost")
```

```{r, eval = FALSE, out.width = "60%"}
# plot an interactive map of the planning unit cost data
mapview(pu_data, zcol = "cost")
```

**Question**: How many planning units are in the planning unit data?\
**Answer**: There are `r nrow(pu_data)` planning units in the unit data.\

**Question**: What is the highest cost value?\
**Answer**: The highest cost value is \$ `r round(max(pu_data$cost), 2)` in Australian dollars.\

**Question**: Is there a spatial pattern in the planning unit cost values?\
**Answer**: Yes, the planning units with the highest cost tend to be in the northern most central part of Tasmania, followed by the western half of the region.\

## Vegetation data

```{r, out.width = "60%"}
# print a short summary of the data
print(veg_data)

# plot a map of the 20th vegetation class
plot(veg_data[[20]])
```

```{r, eval = FALSE, out.width = "60%"}
# plot an interactive map of the 20th vegetation class
mapview(veg_data[[20]])
```

```{r}
# print number of rows in the data
nrow(veg_data)
```

```{r}
# print number of columns  in the data
ncol(veg_data)
```

```{r}
# print number of cells in the data
ncell(veg_data)
```

```{r}
# print number of layers in the data
nlayers(veg_data)
```

```{r}
# print  resolution on the x-axis
xres(veg_data)
```

```{r}
# print resolution on the y-axis
yres(veg_data)
```

```{r}
# print spatial extent of the grid, i.e. coordinates for corners
extent(veg_data)
```

```{r}
# print the coordinate reference system
print(veg_data@crs)
```

```{r}
# print a summary of the first layer in the stack
print(veg_data[[1]])
```

```{r}
# print the value in the 800th cell in the first layer of the stack
print(veg_data[[1]][800])
```

```{r}
# print the value of the cell located in the 30th row and the 60th column of
# the first layer
print(veg_data[[1]][30, 60])
```

```{r}
# calculate the sum of all the cell values in the first layer
cellStats(veg_data[[1]], "sum")
```

```{r}
# calculate the maximum value of all the cell values in the first layer
cellStats(veg_data[[1]], "max")
```

```{r}
# calculate the minimum value of all the cell values in the first layer
cellStats(veg_data[[1]], "min")
```

```{r}
# calculate the mean value of all the cell values in the first layer
cellStats(veg_data[[1]], "mean")
```

**Question**: What part of the study area is the 13th vegetation class found in (hint: make a map)? For instance, is it in the south-eastern part of the study area?\
**Answer**: The 13th vegetation class is found in the eastern part of the studio area with the most density in the north-eastern part of the study area.\
```{r, out.width = "60%"}
# plot a map of the 13th vegetation class
plot(veg_data[[13]])
```

**Question**: What proportion of cells contain the 12th vegetation class?\
```{r}
veg_data_freq12 <- freq(veg_data[[12]])
```
**Answer**: Of the total cells in the raster (`r ncell(veg_data[[12]])`), `r veg_data_freq12[[2, 2]]` cells contain the 12th vegetation class, making the proportion of the cells that contain the vegetation class `r round(veg_data_freq12[[2, 2]] / ncell(veg_data[[12]]), 4) * 100`%.\

**Question**: Which vegetation class is the most abundant (i.e. present in the greatest number of cells)?\
```{r}
prop_df <- data.frame()

for (i in 1:nlayers(veg_data)){
 prop <- freq(veg_data[[i]])[[2, 2]] / ncell(veg_data)
 prop_df <- rbind(prop_df, prop)
}

prop_df <- rownames_to_column(prop_df) %>% 
  rename(proportion = X0.000317970971120754)

prop_highest <- prop_df %>% filter(proportion == max(proportion))
```

**Answer**: The vegetation class that is most abundant is `r prop_highest[[1]]`.\


# Gap analysis

## Feature abundance

```{r}
# create `prioritizr` problem with only the data
p0 <- problem(pu_data, veg_data, cost_column = "cost")

# print empty problem,
# we can see that only the cost and feature data are defined
print(p0)

# calculate amount of each feature in each planning unit
abundance_data <- feature_abundances(p0)

# print abundance data
print(abundance_data)
```

```{r}
# note that only the first ten rows are printed,
# this is because the abundance_data object is a tibble (i.e. tbl_df) object
# and not a standard data.frame object
print(class(abundance_data))
```

```{r}
# we can print all of the rows in abundance_data like this
print(abundance_data, n = Inf)
```

```{r}
# add new column with feature abundances in km^2
abundance_data$absolute_abundance_km2 <-
  (abundance_data$absolute_abundance * prod(res(veg_data))) %>%
  set_units(m^2) %>%
  set_units(km^2)

# print abundance data
print(abundance_data)
```

```{r}
# calculate the average abundance of the features
mean(abundance_data$absolute_abundance_km2)
```

```{r, out.width = "60%"}
# plot histogram of the features' abundances
hist(abundance_data$absolute_abundance_km2, main = "Feature abundances")
```

```{r}
# find the abundance of the feature with the largest abundance
max(abundance_data$absolute_abundance_km2)
```

```{r}
# find the name of the feature with the largest abundance
abundance_data$feature[which.max(abundance_data$absolute_abundance_km2)]
```

**Question**: What is the median abundance of the features?\
**Answer**: The median abundance from the abundance dataset is `r round(median(abundance_data$absolute_abundance_km2), 2)` $km^2$.\

**Question**: What is the name of the feature with smallest abundance?\
**Answer**: The feature with the smallest abundance is `r abundance_data$feature[which.min(abundance_data$absolute_abundance_km2)]`.\

**Question**: How many features have a total abundance greater than 100 km^2?\
**Answer**: There are `r nrow(abundance_data %>% filter(absolute_abundance_km2 > set_units(100, km^2)))` features with a total abundance greater than 100 $km^2$.\

## Feature representation

```{r}
# create column in planning unit data with binary values (zeros and ones)
# indicating if a planning unit is covered by protected areas or not
pu_data$pa_status <- as.numeric(pu_data$locked_in)

# calculate feature representation by protected areas
repr_data <- eval_feature_representation_summary(p0, pu_data[, "pa_status"])

# print feature representation data
print(repr_data)
```

```{r}
# add new column with the areas represented in km^2
repr_data$absolute_held_km2 <-
  (repr_data$absolute_held * prod(res(veg_data))) %>%
  set_units(m^2) %>%
  set_units(km^2)

# print representation data
print(repr_data)
```

**Question**: What is the average proportion of the features held in protected areas?\
**Answer**: The average proportion of the features held in protected areas is `r round(mean(repr_data$relative_held), 4)`.`\

**Question**: If we set a target of 10% coverage by protected areas, how many features fail to meet this target?\
**Answer**: There are `r nrow(repr_data %>% filter(relative_held > 0.1))` features that fail to meet the target of 10% coverage by protected areas.\

**Question**: If we set a target of 20% coverage by protected areas, how many features fail to meet this target?\
**Answer**: There are `r nrow(repr_data %>% filter(relative_held > 0.2))` features that fail to meet the target of 10% coverage by protected areas.\

**Question**: Is there a relationship between the total abundance of a feature and how well it is represented by protected areas?\
```{r, out.width = "60%"}
plot(abundance_data$absolute_abundance ~ repr_data$relative_held)
```

**Answer**: There does not appear to be a relationship between the total abundance of a feature and how well it is represented by protected areas.\


# Spatial prioritizations

## Starting out simple

```{r, out.width = "65%"}
# print planning unit data
print(pu_data)
```

```{r}
# make prioritization problem
p1_rds <- file.path(dir_data, "p1.rds")
if (!file.exists(p1_rds)){
  p1 <- problem(pu_data, veg_data, cost_column = "cost") %>%
        add_min_set_objective() %>%
        add_relative_targets(0.05) %>% # 5% representation targets
        add_binary_decisions() %>%
        add_lpsymphony_solver()
  saveRDS(p1, p1_rds)
}
p1 <- readRDS(p1_rds)

# print problem
print(p1)

# solve problem
s1 <- solve(p1)

# print solution, the solution_1 column contains the solution values
# indicating if a planning unit is (1) selected or (0) not
print(s1)
```

```{r}
# calculate number of planning units selected in the prioritization
eval_n_summary(p1, s1[, "solution_1"])
```

```{r}
# calculate total cost of the prioritization
eval_cost_summary(p1, s1[, "solution_1"])
```

```{r}
# plot solution
# selected = green, not selected = grey
spplot(s1, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s1",
       colorkey = FALSE)
```

**Question**: How many planing units were selected in the prioritization? What proportion of planning units were selected in the prioritization?\
**Answer**: There were `r eval_n_summary(p1, s1[, "solution_1"])[[2]]` planning units selected in the prioritization. There are a total of `r length(s1$id)` planning units, this means that `r (eval_n_summary(p1, s1[, "solution_1"])[[2]] / length(s1$id)) * 100`% of the planning units were selected.\

**Question**: Is there a pattern in the spatial distribution of the priority areas?\
**Answer**: The priority areas appear to be spread out fairly evenly throughout the region with no spatial pattern.\

**Question**: Can you verify that all of the targets were met in the prioritization (hint: `eval_feature_representation_summary(p1, s1[, "solution_1"])`)?\
**Answer**: \

## Adding complexity

```{r, out.width = "65%"}
# plot locked_in data
# TRUE = blue, FALSE = grey
spplot(pu_data, "locked_in", col.regions = c("grey80", "darkblue"),
       main = "locked_in", colorkey = FALSE)
```

```{r, out.width = "65%"}
# make prioritization problem
p2_rds <- file.path(dir_data, "p2.rds")
if (!file.exists(p2_rds)){
  p2 <- problem(pu_data, veg_data, cost_column = "cost") %>%
      add_min_set_objective() %>%
      add_relative_targets(0.05) %>%
      add_locked_in_constraints("locked_in") %>%
      add_binary_decisions() %>%
      add_lpsymphony_solver()
  saveRDS(p2, p2_rds)
}
p2 <- readRDS(p2_rds)

# print problem
print(p2)

# solve problem
s2 <- solve(p2)
print(s2)
```

```{r, out.width = "65%"}
# plot solution
# selected = green, not selected = grey
spplot(s2, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s2",
       colorkey = FALSE)
```

Let's pretend that we talked to an expert on the vegetation communities in our study system and they recommended that a 10% target was needed for each vegetation class. So, equipped with this information, let's set the targets to 10%.

```{r, out.width = "65%"}
# make prioritization problem
p3_rds <- file.path(dir_data, "p3.rds")
if (!file.exists(p3_rds)){
  p3 <- problem(pu_data, veg_data, cost_column = "cost") %>%
    add_min_set_objective() %>%
    add_relative_targets(0.1) %>%
    add_locked_in_constraints("locked_in") %>%
    add_binary_decisions() %>%
    add_lpsymphony_solver()
  saveRDS(p3, p3_rds)
}
p3 <- readRDS(p3_rds)

# print problem
print(p3)

# solve problem
s3 <- solve(p3)
print(s3)
```

```{r, out.width = "65%"}
# plot solution
# selected = green, not selected = grey
spplot(s3, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s3",
       colorkey = FALSE)
```

Next, let's lock out highly degraded areas. Similar to before, this information is present in our planning unit data so we can use the `locked_out` column name to achieve this.

```{r, out.width = "65%"}
# plot locked_out data
# TRUE = red, FALSE = grey
spplot(pu_data, "locked_out", col.regions = c("grey80", "darkred"),
       main = "locked_out", colorkey = FALSE)
```

```{r}
# make prioritization problem
p4_rds <- file.path(dir_data, "p4.rds")
if (!file.exists(p4_rds)){
  p4 <- problem(pu_data, veg_data, cost_column = "cost") %>%
    add_min_set_objective() %>%
    add_relative_targets(0.1) %>%
    add_locked_in_constraints("locked_in") %>%
    add_locked_out_constraints("locked_out") %>%
    add_binary_decisions() %>%
    add_lpsymphony_solver()
  saveRDS(p4, p4_rds)
}
p4 <- readRDS(p4_rds)

# print problem
print(p4)

# solve problem
s4 <- solve(p4)
print(s4)
```

```{r, out.width = "65%"}
# plot solution
# selected = green, not selected = grey
spplot(s4, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s4",
       colorkey = FALSE)
```

```{r, include=FALSE}
assert_that(
  !identical(s3$solution_1, s4$solution_1),
  eval_cost_summary(p3, s3[, "solution_1"])$cost <
    eval_cost_summary(p4, s4[, "solution_1"])$cost)
```

**Question**: What is the cost of the planning units selected in `s2`, `s3`, and `s4`?\
**Answer**: The cost of the planning units selected in `s2` is \$`r round(eval_cost_summary(p1, s2[, "solution_1"])[[2]], 2)`. The cost of the planning units selected in `s3` is \$`r round(eval_cost_summary(p1, s3[, "solution_1"])[[2]], 2)`. And the costs of the planning units selected in `s4` is \$`r round(eval_cost_summary(p1, s4[, "solution_1"])[[2]], 2)`.\

**Question**: How many planning units are in `s2`, `s3`, and `s4`?\
**Answer**: \

**Question**: Do the solutions with more planning units have a greater cost? Why (or why not)?\
**Answer**: \

**Question**: Why does the first solution (`s1`) cost less than the second solution with protected areas locked into the solution (`s2`)?\
**Answer**: \

**Question**: Why does the third solution (`s3`) cost less than the fourth solution solution with highly degraded areas locked out (`s4`)?\
**Answer**: \

## Penalizing fragmentation

```{r, out.width = "65%"}
# make prioritization problem
p5_rds <- file.path(dir_data, "p5.rds")
if (!file.exists(p5_rds)){
  p5 <- problem(pu_data, veg_data, cost_column = "cost") %>%
    add_min_set_objective() %>%
    add_boundary_penalties(penalty = 0.001) %>%
    add_relative_targets(0.1) %>%
    add_locked_in_constraints("locked_in") %>%
    add_locked_out_constraints("locked_out") %>%
    add_binary_decisions() %>%
    add_lpsymphony_solver()
  saveRDS(p5, p5_rds)
}
p5 <- readRDS(p5_rds)

# print problem
print(p5)

# solve problem,
# note this will take a bit longer than the previous runs
s5 <- solve(p5)

# print solution
print(s5)

# plot solution
# selected = green, not selected = grey
spplot(s5, "solution_1", col.regions = c("grey80", "darkgreen"), main = "s5",
       colorkey = FALSE)
```

```{r, include=FALSE}
assert_that(
  !identical(s5$solution_1, s4$solution_1),
  eval_cost_summary(p4, s4[, "solution_1"])$cost <
    eval_cost_summary(p5, s5[, "solution_1"])$cost)
```

**Question**: What is the cost the fourth (`s4`) and fifth (`s5`) solutions? Why does the fifth solution (`s5`) cost more than the fourth (`s4`) solution?\
**Answer**: \

**Question**: Try setting the penalty value to 0.000000001 (i.e. 1e-9) instead of 0.001. What is the cost of the solution now? Is it different from the fourth solution (`s4`) (hint: try plotting the solutions to visualize them)? Is this is a useful penalty value? Why (or why not)?\
**Answer**: \

**Question**: \
**Answer**: Try setting the penalty value to 0.5. What is the cost of the solution now? Is it different from the fourth solution (`s4`) (hint: try plotting the solutions to visualize them)? Is this a useful penalty value? Why (or why not)?\



